---
title: "Summarizing tree samples with ConsTree"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
vignette: >
  %\VignetteIndexEntry{Summarizing tree samples with ConsTree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 2.4, fig.align = "center")
```

'ConsTree' condenses a collection of phylogenetic trees -- a bootstrap or
Bayesian posterior sample, say -- into a single summary tree.  The methods fall
into two families: those that select groupings (splits or clusters) by some
voting rule, and those that summarize the trees through a distance or
treespace criterion.

```{r load, message = FALSE}
library("ConsTree")
library("TreeTools", quietly = TRUE)
```

## Split-selection methods

The split-selection methods differ only in *which* groupings they keep, so they
form a nested sequence of increasing resolution.  To see this, take seven trees
that share a backbone but disagree on the placement of a few leaves:

```{r exemplar}
trees <- ape::read.tree(text = c(
  "((((((t1,t3),t2),t4),(t5,t6)),t7),(t8,t9));",
  "((((((t1,t3),(t5,t6)),t4),t2),t7),(t8,t9));",
  "((((((t1,t2),t3),t4),(t5,t6)),t7),(t8,t9));",
  "(((((t1,(t2,t3)),(t5,t6)),t4),t7),(t8,t9));",
  "((((((t1,t2),t3),t4),(t7,(t5,t6))),t9),t8);",
  "((((((t1,t3),t2),(t5,t6)),t4),t7),(t8,t9));",
  "((((t1,((t2,t3),(t5,t6))),t4),(t8,t9)),t7);"))
```

A consistent colour marks each leaf, so the eye can follow a leaf from one tree
to the next.

```{r palette}
nTip <- 9
leafCol <- setNames(hcl.colors(nTip + 1), TipLabels(nTip + 1))
plotCons <- function(tree, main = "") {
  plot(tree, tip.color = leafCol[tree$tip.label], main = main,
       font = 2, cex = 1, edge.width = 1.5)
}
```

```{r plot-inputs, fig.height = 4.6}
oldPar <- par(mfrow = c(2, 4), mar = c(0.5, 0.5, 1.5, 0.5))
for (i in seq_along(trees)) plotCons(trees[[i]], main = paste("Tree", i))
par(oldPar)
```

Each method retains a superset of the groupings kept by the one before it:

```{r resolution}
data.frame(
  method  = c("Strict", "Majority", "Frequency", "Greedy"),
  splits  = c(NSplits(Strict(trees)),    NSplits(Majority(trees)),
              NSplits(Frequency(trees)),  NSplits(Greedy(trees)))
)
```

```{r plot-gradient, fig.height = 5}
oldPar <- par(mfrow = c(2, 2), mar = c(0.5, 0.5, 1.5, 0.5))
plotCons(Strict(trees),    "Strict")
plotCons(Majority(trees),  "Majority-rule")
plotCons(Frequency(trees), "Frequency difference")
plotCons(Greedy(trees),    "Greedy (extended majority)")
par(oldPar)
```

`Strict()` keeps only the two groupings present in every tree; `Majority()` adds
those in more than half; `Frequency()` keeps a grouping that beats every
grouping it conflicts with; and `Greedy()` adds compatible groupings most
frequent first, giving the most resolved summary.

Two further rules apply different conflict criteria rather than a frequency
threshold.  `Loose()` (the semi-strict, or combinable-component, consensus)
keeps every grouping that *no* tree contradicts, and `MajorityPlus()` keeps a
grouping displayed by more trees than contradict it.  Because the default
`Majority()` threshold (`p = 0.5`) admits a grouping seen in exactly half the
trees, even when the other half contradict it, `Majority()` is *not* always a
subset of `MajorityPlus()`.

## Rooted methods

`Adams()`, `RStar()` and `Local()` treat the input as **rooted** and reason
about clusters and rooted triplets rather than unrooted splits.  They can
therefore recover structure that the unrooted strict consensus collapses:

```{r rooted}
rTrees <- ape::read.tree(text = c(
  "((((t1,t2),t3),(((t4,t6),t9),t5)),t7);",
  "(((((t1,t2),t3),(t4,(t6,t9))),t5),t7);",
  "(((((t1,t2),t3),((t4,t9),t6)),t5),t7);",
  "((((((t1,t2),t9),(t4,t6)),t3),t5),t7);",
  "((((((t1,t2),t3),t9),(t4,t6)),t5),t7);",
  "(((((t1,t2),t9),(t3,(t4,t6))),t5),t7);"))
```

```{r plot-rooted-in, fig.height = 4.6}
oldPar <- par(mfrow = c(2, 3), mar = c(0.5, 0.5, 1.5, 0.5))
for (i in seq_along(rTrees)) plotCons(rTrees[[i]], main = paste("Tree", i))
par(oldPar)
```

```{r rooted-splits}
data.frame(
  method = c("Strict", "Adams", "RStar", "Local"),
  splits = c(NSplits(Strict(rTrees)), NSplits(Adams(rTrees)),
             NSplits(RStar(rTrees)),  NSplits(Local(rTrees)))
)
```

The unrooted strict consensus keeps a single grouping, but the rooted methods
recover more:

```{r plot-rooted-out, fig.height = 5}
oldPar <- par(mfrow = c(2, 2), mar = c(0.5, 0.5, 1.5, 0.5))
plotCons(Strict(rTrees), "Strict")
plotCons(Adams(rTrees),  "Adams")
plotCons(RStar(rTrees),  "RStar")
plotCons(Local(rTrees),  "Local")
par(oldPar)
```

`Adams()` may introduce groupings present in no input tree; `RStar()` keeps each
rooted triplet that wins a plurality over both alternatives; and `Local()`
returns the minimum local consensus of the shared triplets (limited to 20
leaves, and best suited to congruent samples).

## Distance and branch-length summaries

A second family ignores grouping frequencies and instead seeks a tree close to
the sample under a chosen criterion.

The majority-rule tree has a tidy optimality property: it is the *median* tree
under the Robinson–Foulds metric -- the tree whose total RF distance to the
sample is smallest.  `Quartet()` is the quartet-distance analogue, seeking (an
approximation to) the tree that minimizes the total quartet distance to the
inputs instead.  Because the quartet distance gives extra weight to deep
branches, the result is often *more* resolved than the majority-rule tree:

```{r quartet}
c(majority = NSplits(Majority(trees)),
  quartet  = NSplits(Quartet(trees)))
```

When the trees carry **branch lengths**, two further summaries become available.
`Average()` returns the tree whose path-length (patristic) distances best match
the average of the input distance matrices, while `BHVMean()` computes the
Fréchet mean in Billera–Holmes–Vogtmann treespace, with branch lengths;
`BHVDistance()`, `BHVPairwiseDistances()` and `BHVVariance()` provide the
underlying geodesic distances and dispersion.

The Fréchet mean only resolves a grouping that the geodesics actually support.
Given a sample of mutually near-random trees, every internal branch of the mean
is dragged to (almost) zero length: the result is formally resolved but plots as
a star, which is useless to look at.  To show the method's behaviour we instead
need trees that share a backbone.  Here four of five trees agree on the
topology, differing only in branch length, while the fifth rearranges one clade:

```{r branch-length, fig.height = 4.6}
blTrees <- ape::read.tree(text = c(
  "(((t1:0.64,t2:0.84):0.52,(t3:0.84,t4:0.87):0.42):0.46,(t5:0.68,t6:0.34):0.70,t7:0.42);",
  "(((t1:0.64,t2:0.40):0.72,(t3:0.87,t4:0.38):0.87):0.41,(t5:0.58,t6:0.63):0.80,t7:0.63);",
  "(((t1:0.53,t2:0.50):0.78,(t3:0.66,t4:0.37):0.66):0.40,(t5:0.65,t6:0.68):0.48,t7:0.61);",
  "(((t1:0.48,t2:0.47):0.31,(t3:0.46,t4:0.73):0.79):0.65,(t5:0.87,t6:0.34):0.84,t7:0.75);",
  "(((t1:0.85,t2:0.47):0.71,(t4:0.72,(t5:0.78,t6:0.87):0.62):0.36):0.42,t3:0.37,t7:0.46);"))
meanTree <- BHVMean(blTrees)
oldPar <- par(mfrow = c(2, 3), mar = c(0.5, 0.5, 1.5, 0.5))
for (i in seq_along(blTrees)) plotCons(blTrees[[i]], main = paste("Tree", i))
plotCons(meanTree, main = "BHV mean")
par(oldPar)
```

```{r bhv-variance}
BHVVariance(blTrees, mean = meanTree)
```

The mean recovers the shared topology, but its internal branches are not simple
arithmetic averages.  The two groupings that every tree displays keep their full
length, whereas the two that the fifth tree contradicts are pulled noticeably
shorter -- the geodesic to a tree that lacks a branch shrinks that branch on the
way, so disagreement is expressed as shortening rather than collapse.

## Choosing a method

As a rule of thumb: `Strict()` and `Loose()` when you want only well-supported
groupings; `Majority()` for the familiar bootstrap summary; `Frequency()`,
`Greedy()` or `Quartet()` for a more resolved picture; the rooted methods when a
root is meaningful; and `Average()` or `BHVMean()` when branch lengths matter.

Where no constructed consensus is wanted, the
['TreeDist'](https://ms609.github.io/TreeDist/) package offers a complementary
approach: it can return the sampled tree with the lowest median
clustering-information distance to the rest -- a single *representative* of the
sample rather than a summary built from its parts.  And the
['Rogue'](https://ms609.github.io/Rogue/) package can first strip unstable
('rogue') leaves, often sharpening any of the consensus trees above.
